---
title: "Data managment"
output:
  html_document:
    df_print: paged
---
# Coordinating information from diverse tables
```{r}
# Load sample data
suppressMessages(library(GSE5859Subset))
data(GSE5859Subset)
```

Upon attachment and loading of package data, we have three data elements:

* geneAnnotation
* geneExpression
* sampleInfo

```{r}
# Check elements
geneExpression[1:2, 1:5]
geneAnnotation[1, ]
sampleInfo[1, ]
```

```{r}
# Integrate data
cbind(sampleInfo[1:3, ], colnames(geneExpression)[1:3], t(geneExpression)[1:3, 1:4])
```

<br>
<br>

# ExpressionSet
```{r}
# Import packages
suppressMessages(library(Biobase))
```

```{r}
# Bind the previous tables in an ExpressionSet
rownames(sampleInfo) = sampleInfo$filename
rownames(geneAnnotation) = geneAnnotation$PROBEID
es5859 = ExpressionSet(assayData = geneExpression)
pData(es5859) = sampleInfo
fData(es5859) = geneAnnotation
```

```{r}
# Subsetting data from an ExpressionSet
es5859_Y = es5859[which(fData(es5859)$CHR == "chrY"), ]
```

<br>

#### Annotation in ExpressionSet
*annotation* argument in an ExpressionSet points to a character describing the platform on which the samples were assayed. This is often the name of a Bioconductor chip annotation package, which facilitated down-stream analysis.

```{r}
# Set annotation for the ExpressionSet
annotation(es5859) = "hgfocus.db"
```

<br>

#### ExperimentData in ExpressionSet
*experimentData* argument in an ExpressionSet points to an optional MIAME (Minimum Information About a Microarray Experiment) instance with meta-data (e.g., the lab and resulting publications from the analysis) about the experiment.

```{r}
# Set experimentData for the ExpressionSet
suppressMessages(library(annotate))
experimentData(es5859) = pmid2MIAME("17206142")
```

```{r}
# Check experimentData
experimentData(es5859)
```

```{r}
# Check abstract
abstract(es5859)
```

<br>
<br>

# Expression array archives
Expression array archive in the US:

* Gene Expression Omnibus (GEO)
* Package: *GEOquery*

Expression array archive in Europe:

* European Molecular Biology Laboratories (EMBL)
* Package: *ArrayExpress*

<br>

#### Obtaining the ExpressionSet for a GEO series
```{r}
suppressMessages(library(GEOquery))
# glioMA = getGEO("GSE78703")[[1]]
```

<br>

#### Obtaining the ExpressionSet for a EMBL series
```{r}
suppressMessages(library(ArrayExpress))
# glioMA = getAE("E-MTAB-5797")
```

<br>
<br>

# SummarizedExperiment
*colData()*: get the sample-level data

```{r}
# Load sample data (RNA-seq study)
suppressMessages(library(airway))
suppressMessages(library(SummarizedExperiment))
data(airway)
```

```{r}
# Get metadata about the experiment
metadata(airway)
```

```{r}
# Get the first five feature names
rownames(airway)[1:5]
```

```{r}
# Get the first five exon coordinates for a specific gene
rowRanges(airway)$ENSG00000172057[1:5]
```

```{r}
# Get the sample-level data
colData(airway)
```

```{r}
# Check for the existence of overlapping regions in the exon coordinates of one specific gene
reduce(rowRanges(airway)$ENSG00000172057)
```

```{r}
# Summarize data for the sample data
table(airway$dex)
```
