---
title: "GRanges"
output:
  html_document:
    df_print: paged
---

```{r}
# Import packages
library(GenomicRanges)
```

```{r}
# Create an example gene
gen1 <- GRanges("chr1",IRanges(1000001,1001000),strand="+")
```

```{r}
# Get gene spec
## Start
start(gen1)
## End
end(gen1)
## Width
width(gen1)
## Strand
strand(gen1)
## The 'metadata columns', any information stored alongside each range
mcols(gen1)
## IRanges
ranges(gen1)
## The chromosomes for each ranges
seqnames(gen1)
## The possible chromosomes
seqlevels(gen1)
## The lengths for each chromosome
seqlengths(gen1)
```

<br>

### Intra-range methods
Affects ranges independently

function    |   description
---         |   ---
shift       |   moves left/right
narrow      |   narrows by relative position within range
resize      |   resizes to width, fixing start for +, end for -
flank       |   returns flanking ranges to the left +, or right -
promoters   |   similar to flank
restrict    |   restricts ranges to a start and end position
trim        |   trims out of bound ranges
+/-         |   expands/contracts by adding/subtracting fixed amount
*           |   zooms in (positive) or out (negative) by multiples

<br>

### Inter-range methods

Affects ranges as a group

function    |   description
---         |   ---
range       |   one range, leftmost start to rightmost end
reduce      |   cover all positions with only one range
gaps        |   uncovered positions within range
disjoin     |   breaks into discrete ranges based on original starts/ends