---
title: "Inference"
output:
  html_document:
    df_print: paged
---
# t-test

```{r}
# Import packages and the previous data for individuals
suppressMessages(library(Biobase))
suppressMessages(library(maPooling))
data(maPooling)
pd = pData(maPooling)
individuals = which(rowSums(pd) == 1)
individuals = individuals[-grep("tr", names(individuals))]
ind_y = exprs(maPooling)[, individuals]
ind_g = factor(as.numeric(grepl("b", names(individuals))))
```

```{r}
# t-test
suppressMessages(library(genefilter))
ind_tt = rowttests(ind_y, ind_g)

# How many are sig at p < 0.01
sum(ind_tt$p.value < 0.01)
```

<br>
<br>

# Multiple comparisons

```{r}
# Address the issue using the q-value
suppressMessages(library(qvalue))
qvals = qvalue(ind_tt$p.value)$qvalue
sum(qvals < 0.01)
```

<br>
<br>

# The limma hierarchical model

#### Simple t-tests

```{r}
# Import data
suppressMessages(library(SpikeInSubset))
data(rma95)
```

```{r}
# Check pheno data for the first 5 mRNAs
pData(rma95)[, 1:5]

# Check seq data for five mRNAs
exprs(rma95)[655:659, ]
```

```{r}
# Simple t-tests
suppressMessages(library(genefilter))
fac <- factor(rep(1:2, each = 3))
rtt <- rowttests(exprs(rma95), fac)
```

```{r}
# Define colors depending on whether the p-value is small, the absolute difference in means is large, and whether the feature is a spike-in value
mask <- with(rtt, abs(dm) < 0.2 & p.value < 0.01)
mask[140:160]
spike <- rownames(rma95) %in% colnames(pData(rma95))
spike[20:40]

# The spike-in genes are in blue, while the red points indicate genes which have small p-values but also small differences in means.
cols <- ifelse(mask, "red", ifelse(spike, "dodgerblue", "black"))
cols[20:30]

# Plot p-values against DM
with(
  rtt,
  plot(
    -dm,
    -log10(p.value),
    cex = .8,
    pch = 16,
    xlim = c(-1, 1),
    ylim = c(0, 5),
    xlab = "difference in means",
    col = cols
  )
)
abline(h = 2, v = c(-.2, .2), lty = 2)

# Plot p-values against SD
rtt$s <-
  apply(exprs(rma95), 1, function(row)
    sqrt(.5 * (var(row[1:3]) + var(row[4:6]))))
with(
  rtt,
  plot(
    s,
    -log10(p.value),
    cex = .8,
    pch = 16,
    log = "x",
    xlab = "estimate of standard deviation",
    col = cols
  )
)
```

<br>

#### limma steps

```{r}
suppressMessages(library(limma))

# Step 1: least squares estimates
fit <- lmFit(rma95, design = model.matrix( ~ fac))

# Step 2: moderate the t statistics
fit <- eBayes(fit)

# Step 3: report
topTable(fit, coef = 2)
```

**Note:** *topTable* will return the top genes ranked by whichever value you define.

**Note:** The column *adj.P.Val* includes the adjusted p-values (using the method of Benjamini-Hochberg) for each gene.

```{r}
# Plot p-values against DM
limmares <-
  data.frame(dm = coef(fit)[, "fac2"], p.value = fit$p.value[, "fac2"])
with(
  limmares,
  plot(
    dm,
    -log10(p.value),
    cex = .8,
    pch = 16,
    col = cols,
    xlab = "difference in means",
    xlim = c(-1, 1),
    ylim = c(0, 5)
  )
)
abline(h = 2, v = c(-.2, .2), lty = 2)
```

Note that the red points are now all under the line where -log10(p.value) is equal to 2. Also, the blue points which represent real differences have p-values which are even higher than before.
