---
title: "Multi-Omic Data Integration"
output: 
  pdf_document: 
    toc: yes
    number_sections: yes
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
rm(list = ls())
```

# Integrative analysis examples
### TF binding and expression co-regulation in yeast
There are two key experimental components to this example:

* Protein binding patterns: based on ChIP-chip experiments, we can determine the gene promoter regions to which transcription factors bind.
* Expression patterns: based on timed observations of gene expression in a yeast colony we can identify times at which groups of genes reach maximal expression.

```{r}
# Import data
library(harbChIP)
data(harbChIP)

# Check paper's abstract
abstract(harbChIP)

# Check expression data
exprs(harbChIP)[1:5, 1:5]

# Check feature data
fData(harbChIP)[1:4, 1:9]

# Check pheno data (it's TFs' names)
sampleNames(harbChIP)[1:3]
```

Now let’s find MBP1 transcription factor and assess the distribution of reported binding affinity measures.

```{r}
qnorm(exprs(harbChIP)[, "MBP1"])
```

The shape of the qq-normal plot is indicative of a strong departure from Gaussianity in the distribution of binding scores. Now let's focus on the top five genes:
```{r}
topb = featureNames(harbChIP)[order(exprs(harbChIP)[, "MBP1"], decreasing = TRUE)[1:5]]

# Map to reference
library(org.Sc.sgd.db)
select(org.Sc.sgd.db,
       keys = topb,
       keytype = "ORF",
       columns = "COMMON")
```

